Who we are affects how we vote

### 2019-05-08 elections_conversation.tex
Finished for now

A simple way to measure voting patterns is to consider the two-party preferred (2PP) vote, which is based on the tally of preferences for the Labor and Liberal candidates, ignoring all other candidates. By convention, this is recorded as a percentage preference in favour of the Liberal party --- for example, a 2PP value of 45\% indicates that 45\% of voters ranked the Liberal candidate higher than the Labor candidate, while the remaining 55\% ranked them in the reverse order.

@@

Each election was modelled separately, to allow us to see any changes over time, and to account for changing electorate boundaries. In this article, we highlight the variables with the strongest relationship to the two-party preferred vote, or which have had substantial changes over time. The full analysis is available at \url{https://robjhyndman.com/publications/elections/}.

@@

This shows that the baseline of party preference has varied over the elections, with the biggest swing occurring in the 2007 election where the mean electorate shifted more than five percentage points in favour of the Labor party. The dots represent the estimated 2PP value, and the lines indicate a 95\% confidence interval providing a guide to the uncertainty in the estimate. Where the vertical lines corresponding to each election cross the 50\% horizontal line, the 2PP vote for an average electorate is not statistically distinguishable from 50\% for that election.

@@

Regions comprising more older people are often believed to be more conservative, and indeed we found that electorates with a higher median age are more likely to support the Liberal party --- although this effect is significant only in 2004 and 2007.

@@

This graph shows the effect of median age on the 2PP vote. The vertical scale can be interpreted as a score measuring how much the 2PP vote differs between an average electorate, and an electorate with median age in the top 10\% of all electorates (with all other socio-demographic variables unchanged). The reverse is true for young electorates --- the score can be interpreted as how much the 2PP vote would be reduced if an average electorate kept all socio-demographic variables fixed but median age was reduced to be in the bottom 10\% of all eletorates.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates appear more likely to support the Liberal party, as the \texttt{Incomes} factor has a positive effect on Liberal preference. The effect also seems to have been increasing over time.

@@

Unemployment however, is not as influential. In 2001 and 2004, electorates with higher unemployment align with Labor, but over time this shifts towards support for the Liberal party, culminating in a significantly positive (but small) effect in 2016.

@@

Electorates with higher proportions of workers in extractive industries (mining, gas, water, agriculture, waste and electricity) are consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years.

@@

This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production.

Electorates with more workers in transformative industries (construction or manufacturing) are also more likely to support the Liberal party.

@@

The proportion of workers in managerial, administrative, clerical and sales roles is also a significant predictor of two-party preference vote across all six elections, with a higher proportion of people working these jobs increasing Liberal support.

@@

In each of the six elections, electorates with a higher proportion of people that have recently (in the past five years) moved house are more likely to support the Liberal party.

@@

Speaking languages other than English however, appears to have a far stronger effect. Electorates with more diverse languages are associated with higher support for the Liberal party from 2004 onwards, with this effect being significant in 2007, 2010 and 2016.

@@

Of the variables relating to religion, only Judaism shows a consistent effect, with electorates with relatively large Jewish populations more likely to vote Liberal.

@@

It is often suggested that states have systematic differences that cause their electorates to vote differently. We can explore this effect by looking at the difference between the actual 2PP vote in each electorate and what our modelling would predict based only on electorate characteristics. We found a bias toward Labor in the ACT, and to a smaller extent in Tasmania, and a bias toward the Liberals in the Northern Territory. No other states showed a significantly different voting pattern from what you would expect using only socio-demographic information. That suggests that any differences between Queensland and Victoria (for example) are due to different voter socio-demographics.

@@

Based on the distribution of the Cook's distance values, a Cook's distance greater than \(0.1\) is considered to be influential and a potential outlier. The electorate of Sydney (NSW) has a large Cook's distance from 2001 to 2013, due to its diverse population (language, birthplace and religion), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It has remained a strong supporter of the Labor party and the Liberal vote is severely overpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth is found to be an outlier in all but the 2007 election. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup --- implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. Liberal support in Wentworth is underpredicted by the model in each year, and more so with Turnbull as Liberal leader.

Lingiari, an electorate taking up almost all of the Northern Territory, is an outlier in the 2001--2007 elections due to its large Indigenous population, young age profile and low rates of property ownership. Fowler (NSW) has a diverse population with a high proportion of migrants, many Buddhists and Muslims, and has strong Labor support, making it influential in 2001, 2004 and 2010. Other electorates with large Cook's distance are Barton (NSW) and Leichhardt (QLD) in 2016, and Canberra (ACT) in 2007.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This paper explores the effects of electoral socio-demographic characteristics on the two-party preferred vote in the 2001--2016 elections, using information from the corresponding Australian federal elections and Censuses. As a Census does not always occur in the same year as an election, Census data for the 2004--2013 elections are generated by employing a method of spatio-temporal imputation. This imputes electoral socio-demographics for the electoral boundaries in place at the time of the election --- an approach that is distinctly different from previous work on modelling election outcomes, where Census and election data are typically joined without addressing their temporal differences. Before estimating a model, these socio-demographic variables are standardized (to adjust for changing variable scales) and many variables (representing similar information) are combined into factors, resulting in a reduced predictor set. A spatial error model is then estimated for each election, accounting for the inherent spatial structure of the data.

Across the past six elections, most of the socio-demographics that drive the electoral two-party preferred vote are found to remain steady, whilst a few (typically weaker) effects vary over time. Industry and type of work are particularly influential, with energy-related and manufacturing/construction jobs, as well as administrative roles being strongly linked with the Liberal party in all elections. Incomes have a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment shift from weak association with Labor to a significant Liberal effect over the years, and higher education levels are associated with Labor from 2007 (although significant only in 2016). It is also found that electorates with higher household mobility support Liberal, birthplace diversity favours Labor and more de facto relationships align with Labor preference --- although marriages, family and household sizes have no material influence. Furthermore, the neighbourhood (spatial) effects are found to be positive in all elections, although significant only in 2001 and 2016, meaning that in the 2004--2013 elections, electorates effectively voted independently.

The findings in this paper complement the existing literature by modelling temporal trends, which as far as the authors are aware, has not been done previously for Australian elections using a regression framework. It is also the first study to model any Australian election since 2010 using Census information.

Additionally, a key contribution of this research is the wrangling of the raw data and imputed data sets for the 2004, 2007, 2010 and 2013 elections, which have been contributed to the \texttt{eechidna} \texttt{R} package --- providing a rich, accessible data resource for future Australian electoral analysis.

@@

This paper was produced using \texttt{RMarkdown} \autocite{rmarkdown} and \texttt{knitr} \autocite{knitr}. All corresponding code for this paper can be found in the github repository \href{https://github.com/jforbes14/eechidna-paper}{github.com/jforbes14/eechidna-paper}, and the data used is available in the \texttt{eechidna} package \autocite{eechidna}. All raw data was obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government.

\hypertarget{software}{%
\section{Software}\label{software}}

All election and Census datasets, along with electoral maps and more, are available in the \texttt{eechidna} (Exploring Election and Census Highly Informative Data Nationally for Australia) \texttt{R} package, which can be downloaded from CRAN. The \texttt{eechidna} package makes it easy to look at the data from the Australian Federal elections and Censuses that occurred between 2001 and 2016. This study contributed a large revision to the \texttt{eechidna} package, which included the addition of election and Census data for 2001--2010, voting outcomes for polling booths and imputed Census data for election years. For more details on using \texttt{eechidna}, please see the articles (vignettes) on the github page \href{https://ropenscilabs.github.io/eechidna/}{ropenscilabs.github.io/eechidna/}.

The authors would like to sincerely thank Anthony Ebert, Heike Hofmann, Thomas Lumley, Ben Marwick, Carson Sievert, Mingzhu Sun, Dilini Talagala, Nicholas Tierney, Nathaniel Tomasetti, Earo Wang and Fang Zhou, all of whom have contributed to the \texttt{eechidna} package.

>>>>

A simple way to measure voting patterns is to consider the two-party preferred (2PP) vote, which is based on the tally of preferences for the Labor and Coalition candidates, ignoring all other candidates. We express this as a percentage preference in favour of the Coalition --- for example, a 2PP value of 45\% indicates that 45\% of voters ranked the Coalition candidate higher than the Labor candidate, while the remaining 55\% ranked them in the reverse order.

@@

Each election was modelled separately, to allow us to see any changes over time, and to account for changing electorate boundaries. In this article, we highlight the variables with the strongest relationship to the two-party preferred vote, or which have had substantial changes over time.

@@

This shows that the baseline of party preference has varied over the elections, with the biggest swing occurring in the 2007 election where the average electorate shifted more than five percentage points in favour of the Labor party. The dots represent the estimated 2PP value, and the lines indicate a 95\% confidence interval providing a guide to the uncertainty in the estimate. Where the vertical lines corresponding to each election cross the 50\% horizontal line, the 2PP vote for an average electorate is not statistically distinguishable from 50\% for that election.

@@

Regions comprising more older people are often believed to be more conservative, and indeed we found that electorates with a higher median age are more likely to support the Coalition party.

@@

This graph shows the effect of median age on the 2PP vote. The vertical scale can be interpreted as a score measuring how much the 2PP vote differs between an average electorate, and an electorate with median age in the top 10\% of all electorates (with all other socio-demographic variables unchanged). The reverse is true for young electorates --- the score can be interpreted as how much the 2PP vote would be reduced if an average electorate kept all socio-demographic variables fixed but median age was reduced to be in the bottom 10\% of all electorates.

When the uncertainty range covers the horizontal 0 line, the resulting impact can be considered statistically undetectable. Here age is a statistically significant effect only in 2004 and 2007.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates appear more likely to support the Coalition, as the Incomes factor has a positive effect on Coalition preference. The effect also seems to have been increasing over time.

@@

Unemployment however, is not as influential. In 2001 and 2004, electorates with higher unemployment align with Labor, but over time this shifts towards support for the Coalition, culminating in a significantly positive (but small) effect in 2016.

@@

Electorates with higher proportions of workers in extractive industries (mining, gas, water, agriculture, waste and electricity) are consistently linked with higher support for the Coalition, with the magnitude of this effect slightly increasing over the years.

@@

This is unsurprising, as the Coalition has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production.

Electorates with more workers in transformative industries (construction or manufacturing) are also more likely to support the Coalition.

@@

The proportion of workers in managerial, administrative, clerical and sales roles is also a significant predictor of two-party preference vote across all six elections, with a higher proportion of people working these jobs increasing Coalition support.

@@

In each of the six elections, electorates with a higher proportion of people that have recently (in the past five years) moved house are more likely to support the Coalition.

@@

Speaking languages other than English however, appears to have a far stronger effect. Electorates with more diverse languages are associated with higher support for the Coalition from 2004 onwards, with this effect being significant in 2007, 2010 and 2016.

@@

Of the variables relating to religion, only Judaism shows a consistent effect, with electorates with relatively large Jewish populations more likely to vote for the Coalition.

@@

It is often suggested that states have systematic differences that cause their electorates to vote differently. We can explore this effect by looking at the difference between the actual 2PP vote in each electorate and what our modelling would predict based only on electorate characteristics. We found a bias toward Labor in the ACT, and to a smaller extent in Tasmania, and a bias toward the Coalition in the Northern Territory. No other states showed a significantly different voting pattern from what you would expect using only socio-demographic information. That suggests that any differences between Queensland and Victoria (for example) are due to different voter socio-demographics.

@@

By this measure, the top five most surprising results in the last six elections were the following:

\begin{tabular}{llrr}
\toprule
Electorate & Election & Actual 2PP & Predicted 2PP\\
\midrule
Capricornia & 2004 & 44.9 & 60.6\\
Capricornia & 2001 & 43.1 & 58.8\\
Herbert & 2013 & 56.2 & 41.6\\
Wentworth & 2013 & 67.7 & 53.1\\
Herbert & 2016 & 50.0 & 35.8\\
\bottomrule
\end{tabular}

Capricornia is on the Queensland coast, taking in Rockhampton, Yeppoon and parts of Mackay. It was won in 2001 and 2004 by Labour MP Kirsten Livermore, defying the socio-demographic prediction which suggested it would be a safe Coalition seat. She also won it in 2007 and 2010, but with a smaller number of votes, so the result was not so different from predicted. It is currently held by Michelle Landry for the Coalition.

Herbert is another Queensland seat taking in Townsville. It was won in 2013 for the Coalition by Ewen Jones, and in 2016 by Cathy O'Toole for Labor by just 37 votes. The model predicts a much lower vote for the Coalition than tends to happen in practice.

Wentworth probably needs no explanation --- the seat formerly held by Malcolm Turnbull which he won with a big majority in 2013. Its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup --- implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. Liberal support in Wentworth is underpredicted by the model in each year, and more so with Turnbull as Liberal leader.

\hypertarget{true-blue-seats}{%
\section{True Blue Seats}\label{true-blue-seats}}

Another way the model can be used is to identify the seats with the strongest propensity to vote one way or the other, based on their socio-demographics.

The top five most blue-ribbon Coalition seats in the last six elections are listed below. Mitchell takes in the north-western outer suburbs of Sydney, including Baulkham Hills, Kellyville, and parts of Castle Hill. Maranoa occupies a large part of south west Queensland including towns such as Charleville, Kingaroy and Winton. Mallee is, of course, the north west corner of Victoria including Mildura, Swan Hill and Stawell.

\begin{tabular}{llrr}
\toprule
Electorate & Election & Actual 2PP & Predicted 2PP\\
\midrule
Mitchell & 2001 & 71.3 & 76.4\\
Mitchell & 2004 & 70.7 & 75.9\\
Maranoa & 2013 & 72.3 & 75.6\\
Maranoa & 2016 & 67.5 & 73.9\\
Mallee & 2013 & 73.7 & 73.8\\
\bottomrule
\end{tabular}

The top five seats most likely to vote Labour, based on their socio-demographics, are listed below. Batman has since been abolished, but occupied the inner north of Melbourne. Gellibrand takes in the inner western suburbs of Melbourne including Altona and Williamstown, while Fowler is in the western suburbs of Sydney, including Cabramatta and Liverpool.

\begin{tabular}{llrr}
\toprule
Electorate & Election & Actual 2PP & Predicted 2PP\\
\midrule
Batman & 2007 & 24.1 & 22.3\\
Batman & 2010 & 25.1 & 22.5\\
Fowler & 2004 & 28.6 & 23.6\\
Fowler & 2001 & 28.5 & 24.9\\
Gellibrand & 2010 & 26.1 & 25.2\\
\bottomrule
\end{tabular}

@@

This analysis is based on a research paper coauthored by Rob Hyndman and Di Cook from the Department of Econometrics and Business Statistics at the Monash Business School, and Jeremy Forbes, a former Monash University Honours student in econometrics. The full analysis is available at \url{https://robjhyndman.com/publications/elections/}.

The code used for the analysis can be found in the github repository \href{https://github.com/jforbes14/eechidna-paper}{github.com/jforbes14/eechidna-paper}.

All election and Census data since 2001, along with electoral maps and more, are available in the \href{https://cran.r-project.org/package=eechidna}{\texttt{eechidna}} (Exploring Election and Census Highly Informative Data Nationally for Australia) \texttt{R} package, which can be downloaded from CRAN. The eechidna package uses data obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government.
